---
title: "PT/FT Group Differences Analyses - Preliteracy outcomes"
author: "Virginia Marchman"
date: "3/30/2023, updated 4/15/2023"
output:
  html_document: 
   toc: true
   toc_float: true
---

```{r, echo = F}
library(knitr)
opts_chunk$set(echo=TRUE, 
               warning=FALSE, message=FALSE, 
               cache=FALSE)
options(width = 100)
```

# This is the code to generate the analyses in Marchman, V.A., Ashland, M., Loi, E.C., Shannon, K. A., Munevar, M., Fernald, A, & Feldman, H. M. Early language processing efficiency and pre-literacy outcomes in children born full term and preterm


Load libraries and set theme
```{r}
library(tidyverse)
library(effsize)
library(stargazer)
library(psych)
library(powerAnalysis)

theme_set(theme_bw())
```


# Data prepping
### Load data
```{r}

setwd = "G:/Shared drives/Lang Learning Lab Management Archived/Studies Archive/PDEV Series/PDEV/PDEV Papers/PTFT Group Differences"

PTFTlit <- read_csv("PTFT_data_literacy.csv")

PTFTlit <- as.data.frame(PTFTlit)
```

### Check data
```{r}
# Sex - 46 females, 49 males
PTFTlit %>% 
  group_by(Group, Sex) %>% 
  count()
```

### Transform variables (e.g., character to factor)
```{r}
PTFTlit <- PTFTlit %>% 
  mutate(SubjectID = factor(SubjectID),
         Sex = factor(Sex), Group = factor(Group))
```

#Demographic variables
### Descriptive statistics - t-tests by Birth Group - Table 1

```{r}

t.test(GADays ~ Group, data = PTFTlit, var.equal = FALSE)
t.test(BirthWtGm ~ Group, data = PTFTlit, var.equal = FALSE)
t.test(MomEdu ~ Group, data = PTFTlit, var.equal = FALSE)
t.test(HI_SES ~ Group, data = PTFTlit, var.equal = FALSE)
t.test(AgeAtFirstVisit_PDEV18ATL318C ~ Group, data = PTFTlit, var.equal = FALSE)
t.test(AgeAtFirstVisit_PDEV54CTL354C ~ Group, data = PTFTlit, var.equal = FALSE)

```
### Number of trials with known words by birthgroup.
```{r}


PTFTlit %>% group_by(Group) %>%
  summarise(mean_Known = mean(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A, na.rm = T),
            sd_Known = sd(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A, na.rm = T),
            min_Known = min(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A, na.rm = T),
            max_Known = max(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A, na.rm = T))

t.test(PeekVocabChecklist_UnderstandsKnownWords_TL318PDEV18A ~ Group, data = PTFT)

```

### Number of trials for RT and test for differences between birthgroup.
```{r}

PTFTlit %>% group_by(Group) %>%
  summarise(mean_trials = mean(DRTAB183001800knownnumPDEV18ATL318C, na.rm = T), 
            sd_trials = sd(DRTAB183001800knownnumPDEV18ATL318C, na.rm = T),
            min_trials = min(DRTAB183001800knownnumPDEV18ATL318C, na.rm = T), 
            max_trials = max(DRTAB183001800knownnumPDEV18ATL318C, na.rm = T))

t.test(DRTAB183001800knownnumPDEV18ATL318C ~ Group, data = PTFT)

```


#Number of trials for Accuracy and test for differences between birthgroups.
```{r}

PTFTlit %>% group_by(Group) %>%
  summarise(mean_trials = mean(AccAB183001800knownnumPDEV18ATL318C, na.rm = T), 
            sd_trials = sd(AccAB183001800knownnumPDEV18ATL318C, na.rm = T),
            min_trials = min(AccAB183001800knownnumPDEV18ATL318C, na.rm = T), 
            max_trials = max(AccAB183001800knownnumPDEV18ATL318C, na.rm = T))

t.test(AccAB183001800knownnumPDEV18ATL318C ~ Group, data = PTFTlit)

```

###LWL measures at 18 months - Table 2

```{r}

## Accuracy - t-test by group

t.test(AccAB183001800knownPDEV18ATL318C ~ Group, data = PTFTlit)

## RT - t-test by group

t.test(RT18A18C_NO1RT ~ Group, data = PTFTlit)

```

# Child language measures at 4.5 years - Table 2
```{r}

PTFTlit %>% group_by(Group) %>%
  summarise(
        mean_PrintK = mean(PrintKnowledge_PDEV54TL354_TOPEL_standard, na.rm = T), 
        sd_PrintK = sd(PrintKnowledge_PDEV54TL354_TOPEL_standard, na.rm = T),
        mean_Phon = mean(PhonAwareness_PDEV54TL354_TOPEL_standard, na.rm = T), 
        sd_Phon = sd(PhonAwareness_PDEV54TL354_TOPEL_standard, na.rm = T),
        mean_Exp = mean(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard, na.rm = T), 
        sd_Exp = sd(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard, na.rm = T))


# Print Knowledge
t.test(PrintKnowledge_PDEV54TL354_TOPEL_standard ~ Group, data = PTFTlit)

# Phonological Awareness
t.test(PhonAwareness_PDEV54TL354_TOPEL_standard ~ Group, data = PTFTlit)

# CELF
t.test(CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard ~ Group, data = PTFTlit)

```

# Regression models
### Models 1a to 1e - Print Knowledge - Table 3
```{r}
# covariates SES and Age and Group

m1a <- lm(PrintKnowledge_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group, data = PTFTlit)

# covariates + Group + Accuracy
m1b <- lm(PrintKnowledge_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Group + AccAB183001800knownPDEV18ATL318C, data = PTFTlit)

# covariates + Group + Accuracy + Acc x Group
m1c <- lm(PrintKnowledge_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Group * AccAB183001800knownPDEV18ATL318C, data = PTFTlit)

# covariates + Group + RT
m1d <- lm(PrintKnowledge_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group + RT18A18C_NO1RT, data = PTFTlit)

# covariates + Group + RT + RT x Group
m1e <- lm(PrintKnowledge_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group * RT18A18C_NO1RT, data = PTFTlit)

# comparisons
anova(m1a, m1b)
anova(m1b, m1c)
anova(m1a, m1d)
anova(m1d, m1e)

stargazer(m1a, m1b, m1c, m1d, m1e, type = "text",
         star.char = c(".","*","**","***"),
         star.cutoffs = c(.1, .05, .01, .001),
         notes = c(". p<0.1; * p<0.05; ** p<0.01; *** p<0.001"),
         notes.append = F,
         digits = 2, # change to 3 digits to get r2
         dep.var.labels = c("Print Knowledge"),
         covariate.labels=c("Age at test", "SES",
                            "Group",
                            "Acc x Group",
                            "RT",
                            "RT x Group"))

```
### Models 2a to 2e - Phonological Awareness - Table 4
```{r}

# covariates SES and Age and Group

m2a <- lm(PhonAwareness_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group, data = PTFTlit)

# covariates + Group + Accuracy
m2b <- lm(PhonAwareness_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Group + AccAB183001800knownPDEV18ATL318C, data = PTFTlit)

# covariates + Group + Accuracy + Acc x Group
m2c <- lm(PhonAwareness_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C +
            HI_SES + Group * AccAB183001800knownPDEV18ATL318C, data = PTFTlit)

# covariates + Group + RT
m2d <- lm(PhonAwareness_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group + RT18A18C_NO1RT, data = PTFTlit)

# covariates + Group + RT + RT x Group
m2e <- lm(PhonAwareness_PDEV54TL354_TOPEL_standard ~ AgeAtFirstVisit_PDEV54CTL354C + 
            HI_SES + Group * RT18A18C_NO1RT, data = PTFTlit)

# comparisons
anova(m2a, m2b)
anova(m2b, m2c)
anova(m2a, m2d)
anova(m2d, m2e)

stargazer(m2a, m2b, m2c, m2d, m2e, type = "text",
         star.char = c(".","*","**","***"),
         star.cutoffs = c(.1, .05, .01, .001),
         notes = c(". p<0.1; * p<0.05; ** p<0.01; *** p<0.001"),
         notes.append = F,
         digits = 3, # change to 3 digits to get r2
         dep.var.labels = c("Phonological Awareness"),
         covariate.labels=c("Age at test", "SES",
                            "Group",
                            "Acc x Group",
                            "RT",
                            "RT x Group"))
                            
```
# Running mediation using Process in R - Table 5
## Model syntax - model 4
```{r}

# Must run process macro in process.R prior to running code (a user-defined macro)
# The simplest process model syntax is here:
# process (data = my_data_frame, y = "my_DV", x = "my_IV", m ="my_mediator", model = 4)

# with covariates and other parameters specified
# process (data = my_data_frame, y = "my_DV", x = "my_IV", m ="my_mediator", model = 4, effsize =1, total =1, stand =1, cov = c("my_covariate1", "my_covariate2"), boot = 10000 , modelbt = 1, seed = 654321)
```

## Print Knowledge x Accuracy, CELF as mediator, with age at test, SES, and group (must be numeric) as covariates

```{r}
process (data = PTFTlit, y = "PrintKnowledge_PDEV54TL354_TOPEL_standard", x = "AccAB183001800knownPDEV18ATL318C",
  m = "CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard", model = 4, effsize =1, total =1, 
  cov = c("AgeAtFirstVisit_PDEV54CTL354C", "HI_SES", "Groupnum"), boot = 10000 , modelbt = 1, seed = 654321)

```
## Print Knowledge x RT, CELF as mediator, with age at test, SES, and group (must be numeric) as covariates
```{r}

process (data = PTFTlit, y = "PrintKnowledge_PDEV54TL354_TOPEL_standard", x = "RT18A18C_NO1RT",
  m = "CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard", model = 4, effsize =1, total =1, 
  cov = c("AgeAtFirstVisit_PDEV54CTL354C", "HI_SES", "Groupnum"), boot = 10000 , modelbt = 1, seed = 654321)
```
## Phonological Awareness x Accuracy, CELF as mediator, with age at test, SES, and group (must be numeric) as covariates
```{r}

process (data = PTFTlit, y = "PhonAwareness_PDEV54TL354_TOPEL_standard", x = "AccAB183001800knownPDEV18ATL318C",
  m = "CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard", model = 4, effsize =1, total =1, 
  cov = c("AgeAtFirstVisit_PDEV54CTL354C", "HI_SES", "Groupnum"), boot = 10000 , modelbt = 1, seed = 654321)

```
## Phonological Awareness x RT, CELF as mediator, with age at test, SES, and group (must be numeric) as covariates
```{r}

process (data = PTFTlit, y = "PhonAwareness_PDEV54TL354_TOPEL_standard", x = "RT18A18C_NO1RT",
  m = "CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard", model = 4, effsize =1, total =1, 
  cov = c("AgeAtFirstVisit_PDEV54CTL354C", "HI_SES", "Groupnum"), boot = 10000 , modelbt = 1, seed = 654321)

```

# Moderated Mediation - does group moderate the mediation effects?
## Sample syntax: Model 8
```{r}

## PROCESS Model 8
## process (data = my_data_frame, y = "DV", x = "IV", m = "MED", w = "MOD", model = 8, cov = "COV", center = 2, moments = 1, modelbt = 1, boot = 10000, seed = 654321)

```
## Print Knowledge x Accuracy, CELF as mediator, with age at test, SES as covariates, group as moderator
```{r}
process (data = PTFTlit, y = "PrintKnowledge_PDEV54TL354_TOPEL_standard", x = "AccAB183001800knownPDEV18ATL318C",
  m = "CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard", w = "Groupnum", model = 8, effsize =1, total =1, 
  cov = c("AgeAtFirstVisit_PDEV54CTL354C", "HI_SES"), boot = 10000 , modelbt = 1, seed = 654321)

```
## Print Knowledge x RT, CELF as mediator, with age at test, SES as covariates, group as moderator
```{r}

process (data = PTFTlit, y = "PrintKnowledge_PDEV54TL354_TOPEL_standard", x = "RT18A18C_NO1RT",
  m = "CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard", w = "Groupnum", model = 8, effsize =1, total =1, 
  cov = c("AgeAtFirstVisit_PDEV54CTL354C", "HI_SES"), boot = 10000 , modelbt = 1, seed = 654321)
  
```
## Phonological Awareness x Accuracy, CELF as mediator, with age at test, SES as covariates, group as moderator
```{r}

process (data = PTFTlit, y = "PhonAwareness_PDEV54TL354_TOPEL_standard", x = "AccAB183001800knownPDEV18ATL318C",
  m = "CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard", w = "Groupnum", model = 8, effsize =1, total =1, 
  cov = c("AgeAtFirstVisit_PDEV54CTL354C", "HI_SES"), boot = 10000 , modelbt = 1, seed = 654321)

```
## Phonological Awareness x RT, CELF as mediator, with age at test, SES as covariates, group as moderator
```{r}

process (data = PTFTlit, y = "PhonAwareness_PDEV54TL354_TOPEL_standard", x = "RT18A18C_NO1RT",
  m = "CELF_Exp_PDEV54TL354_CELF_ExpressiveLanguage_standard", w = "Groupnum", model = 8, effsize =1, total =1, 
  cov = c("AgeAtFirstVisit_PDEV54CTL354C", "HI_SES"), boot = 10000 , modelbt = 1, seed = 654321)

```
 
